Online supplemental material of the article:\\ \vspace{7mm} 'Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming'\\ \vspace{7mm} \normalsize by

### 2020-04-16 Paper/online_supplemental_material/online_material.tex
edits to online material

\usepackage[numbers,sort&compress]{natbib}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\graphicspath{{./images/}}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}

\usepackage[titletoc]{appendix}

@@

\bibpunct[, ]{[}{]}{,}{}{,}{,}
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}

\usepackage{listings}
\usepackage{slashbox}
\usepackage{caption}
\captionsetup{font=footnotesize}

\usepackage{chngcntr}

\bibliographystyle{authordate1}

\title{Online supplemental material of the article:\\ \vspace{7mm} 'Hilbert space approximate Bayesian Gaussian processes using Stan'\\ \vspace{7mm}
Case Studies}

\author{TODO}

@@

@@

\section{Same-sex marriage case study}\label{ch5_sec_studycaseII}
This data set relates the proportion of support for same-sex marriage to the age. The data consists of 74 observations of the amount of people $y_i$ supporting same-sex marriage from a population $n_i$ per age group $i$ ($i=1,\dots,74$). The observational model is a binomial model with parameters population $n_i$ and probability of supporting same-sex marriage $p_i$ per age group $i$,
%
\begin{equation*}
y_i \sim \text{Binomial}(p_i, n_i).
\end{equation*}

\noindent The population per age group $n_i$ is a known quantity and the goal is to estimate the same-sex support probability $p_i$ or mean number of support people per age group. Probabilities $\bm{p}=(p_1,\dots,p_{74})$ are modeled by a GP function $f:{\rm I\!R} \to {\rm I\!R}$ with a squared exponential covariance function $k$, as a function of age input values $\bm{x}=(x_1,\dots,x_{74})$, and through the {\it logit} function as a link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \mathrm{logit}(f(x_i)) \nonumber \\
f(x) &\sim \mathcal{GP}(0, k(x,x', \theta)).
\end{align*}

\noindent Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,74$. In the HSGP model, the function $f(x)$ is approximated as in equation (7) %(\ref{eq_approxf})
, with the squared exponential spectral density as in equation (1)%(\ref{eq_specdens_inf})
, and eigenvalues $\lambda_j$  and eigenfunctions $\phi_j$ as in equations (5) %(\ref{eq_eigenvalue}) 
and (6)%(\ref{eq_eigenfunction})
.

In order to do model comparison, in addition to the regular GP model and HSGP model, an splines-based model is also fitted using the Thin Plate Regression Splines approach in \cite{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2015package}. A Bayesian approach is used to fit this splines model using the R-package \textit{brms} \citep{burkner2017brms}.

@@

For the HSGP model, different models with different number of basis functions and boundary factor have been fitted. The root mean square errors (RMSE) for every one of these models have been computed against the regular GP model, and plotted as a function of the number of basis functions $m$ and boundary factor $c$ in Figure \ref{ch5_fig13_MSE_train_BF_gaydata}, for sample (left) and test (right) data. The expected patterns of the approximation as a function of the number of basis functions and boundary factor are recognized: as the boundary factor increases, more basis functions are needed.

Figure \ref{ch5_fig14_MSE_train_gaydata} shows the RMSE of the regular GP, HSGP and splines models, computed against the actual data, for training and test data, as a function of the number of basis functions $m$ and boundary factor $c$ for the HSGP model, and  knots for the splines model. We can see how the splines models do not extrapolate data properly.

Figure \ref{ch5_fig14_1_time_gaydata} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the splines model. The computational times is represented in the y-axis which is on a logarithmic scale. The HSGP model is on average roughly 15 times faster than the regular GP and 5 times faster than the spline model, for this particular case and univariate input space. {\color{blue}The increase of computation as function of the number of basis functions in a univariate input space is relatively slight, as can be seen in the figure}.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Same-sex-marriage-data}.

@@

This example consists of a simulated dataset with $n=120$ ($i=1,\dots,n$) single draws from a Gaussian process prior with two input dimensions ($D=2$). A squared exponential covariance function, with hyperparameters marginal variance $\alpha=1$ and lengthscales $\ell_1=0.10$, for the first dimension, and $\ell_2=0.35$, for the second dimension, is used. The corresponding matrix of input values is $X=[\bm{x}_1 \; \bm{x}_2 \; \cdots \; \bm{x}_{120}]^\top \in {\rm I\!R}^{120\times 2}$ with $\bm{x}_i \in \{[-1,1],[-1,1]\} \subset {\rm I\!R}^2$. Gaussian noise $\sigma=0.2$ was added to the GP draws to form the final noisy set of observations $\bm{y} \in {\rm I\!R}^{120}$.

The regular GP model over the outcome variable $\bm{y}$ and matrix of inputs $X \in {\rm I\!R}^{120 \times 2}$, can be written as follows,
%
\begin{eqnarray*}\label{ch5_eq_latentgp_simudata2}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \mathcal{N}(0, \sigma^2  I) \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x}, \bm{x}', \theta)),
\end{split}
\end{eqnarray*}

\noindent where $I$ represents the identity matrix, $\bm{e}$ the noise term, and $\bm{f}=\{f(\bm{x}_i)\}_{i=1}^{120}$ represents the underlying function values to the noisy observations $\bm{y}$. The function $f:{\rm I\!R}^2 \to {\rm I\!R}$ is a GP prior with a multivariate squared exponential covariance function $k$. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $K$, where $K_{rs}=k(\bm{x}_r,\bm{x}_s,\theta)$, with $r,s=1,\dots,120$. The covariance function $k$ depends on the matrix $X$ of risk factors and hyperparameters $\theta=\{\alpha,\ell_1,\ell_2\}$. The hyperparameters $\alpha$, $\ell_1$ and $\ell_2$ represent the marginal variance and lengthscales for first and second dimensions, respectively, of the GP process.

The marginalized form, by integrating out the latent values $\bm{f}$, of the previous latent GP model results:
%
\begin{equation*}\label{ch5_eq_marginalizedgp_simudata2}
\bm{y} \sim \mathcal{N}(0, K + \sigma^2 I ).
\end{equation*}

In the HSGP model with $2$ input dimensions, the latent function $f$, evaluated at input vector $\bm{x} \in {\rm I\!R}^2$, is approximated as in equation (12)% (\ref{eq_approxf_multi})
, with the $2$-dimensional squared exponential spectral density $S$ as in equation (1)% (\ref{eq_specdens_inf})
, and the multivariate eigenfunctions $\phi_j$ and the $2$-vector of eigenvalues $\bm{\lambda}_j$ as in equations (10) %(\ref{eq_eigenfunction_multi})
 and (9)% (\ref{eq_eigenvalue_multi})
 , respectively.

In order to do model comparison, in addition to the regular GP and HSGP models, a two-dimensional splines-based model is also fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2015package}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig15_Posteriors_exII} shows the data-generating GP function, from where the dataset was drawn, and the mean posterior predictive functions of the three models, the regular GP, the HSGP, and the splines, fitted over the dataset. Sample observations are also plotted in the plots as circles. For the HSGP model, $m_1=40$ and $m_2=40$ basis functions for each dimension respectively, were used, which lead to a total of 1600 multivariate basis functions. A boundary factor for each dimension $c_1=1.5$ and $c_2=1.5$ were used. For the splines model, 40 knots in each dimension were used.

@@

In order to assess performance of the models as a function of the number of basis functions and knots, different models with different number of basis functions, for the HSGP model, and different number of knots, for the splines model, have been fitted. In all models, the same number of basis functions and knots per dimension were used. Figure \ref{ch5_fig17_RMSE_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$, and the number of univariate basis functions $m$, for the HSGP model, and knots, for the splines model. From Figures \ref{ch5_fig16_errors_exII} and \ref{ch5_fig17_RMSE_exII}-left, it can be seen a close approximation of the HSGP model to the regular GP model. However, the performance of the splines model is significantly worse. Figure \ref{ch5_fig17_RMSE_exII}-right shows the computational times of the different models as a function of the boundary factor, number of basis functions and knots. Figure \ref{ch5_fig17_RMSE_exII} reveals that choosing the optimal boundary factor allows for less number of basis functions and less computational time. {\color{blue}Even though in a bivariate input space the computation demand increases significantly with the number of dimensions or knots, either the HSGP or spline models work significantly better than regular GP, even for highly wiggly functions where a high number of basis functions or knots are required in the approximations. However, severe difficulties have been found in building the spline model with $50$ knots for this bivariate function}.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_2D-Simulated-data}.

@@

\section{Diabetes case study}\label{ch5_sec_bf_caseV}
The next example presents an epidemiological study of diabetes disease. The study aims to relate the probability of suffering from diabetes to some risk factors. The data contains $n=392$ individuals ($i=1,\dots,n$) from which the binary variable of suffering ($y_i=1$) or not suffering ($y_i=0$) from diabetes have been observed. The matrix $X=[\bm{x}_1\;\bm{x}_2\;\cdots\;\bm{x}_n]^\top \in {\rm I\!R}^{n\times 4}$,  with $\bm{x}_i=(x_{i1},x_{i2},x_{i3},x_{i4}) \in {\rm I\!R}^{4}$, contains the risk factors, \textit{Glucose} ($x_{i1}$), \textit{Pregnancy} ($x_{i2}$), \textit{Age} ($x_{i3}$) and \textit{BMI} ($x_{i4}$), per individual $i$. The observational model is a Bernoulli model with parameter the probability $p_i$ of suffering from diabetes per observation $i$,
%
\begin{equation*}
y_i \sim \text{Bernoulli}(p_i).
\end{equation*}

\noindent The goal is to estimate the probability $p_i$ as a function of the risk factors, which function $f(\cdot):{\rm I\!R}^{4} \to {\rm I\!R}$ is modeled as a Gaussian process with a multivariate squared exponential covariance function $k$ depending on the matrix $X$ of risk factors and hyperparameters $\theta=\{\alpha,\ell\}$, and related to the probabilities $p_i$ through the {\it logit} link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \mathrm{logit}(f(\bm{x}_i)) \nonumber \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x},\bm{x}, \theta).
\end{align*}

\noindent Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}=\{f(\bm{x}_i)\}_{i=1}^n$ are multivariate Gaussian distributed with covariance matrix $K$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,n$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. Notice that a scalar lengthscale is considered in the multivariate covariance function.

@@

In order to do model comparison, in addition to the regular GP and HSGP models, a $D$-dimensional spline-based model is also fitted using a cubic spline basis penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized} and implemented in the R-package \textit{mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms}.

@@

Figure \ref{ch5_fig20_time2D_diabetes} shows the computational times of the different models, regular GP, HSGP and spline, fitted over the dataset, with 2 input dimensions, 3 input dimensions and 4 input dimensions, as a function of the boundary factor $c$ and number of univariate basis functions $m$ and knots. We can appreciate the significant increase of computational time with higher dimensions for the HSGP and spline models. In this sense, choosing optimal values for the number of basis functions and boundary factor, looking at the recommendations and diagnosis provided by Figure 6% \ref{fig6_lscale_vs_J_vs_c}
, can be essential to avoid a excessive computational time, especially in multivariate input spaces. It is interesting to be noticed that considering more than 10 knots per dimension in the spline model with 3D is not allowed for an amount of 392 observations. Similarly, just the computation of the input data for the spline model in a 4D input space is computationally very expensive.

The Stan model codes for the exact GP, the approximate GP and the spline models of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Diabetes-data}.

@@

The next example presents an spatio-temporal classification in land-use of plots between 2006 and 2015 in a part of the territory of Valencia in Spain dedicated to growing citrus fruits. A sampling set consists of $N=200$ plots with known class. The data is recorded in a time series of $T=5$ years (2006, 2008, 2010, 2012, 2015) within the period. The class of each parcel $i$ and time $t$ is stored by a categorical variable $y_{it}$ representing the $K=5$ different possible classes ($k=1,\dots,K$): $k=1$, adult independent citrus fruits; $k=2$, aligned citrus fruits; $k=3$, irregular citrus fruits; $k=4$, abandoned citrus fruits; $k=5$, young citrus fruits.

@@

The principal components (PCs) will be used jointly with the time variable as inputs in the classifying additive HSGP model. Let's denote the matrix $X=[\bm{x}_{11}\, \cdots \, \bm{x}_{it} \, \cdots \, \bm{x}_{NT}]^\top \in {\rm I\!R}^{NT\times D}$, which contains the input vectors $\bm{x}_{it} \in {\rm I\!R}^D$, $D=53$ ($52 \text{ PCs plus time}$) for the spatio-temporal observations (plots $i=1,\dots,200$, and times $t=1,\dots,5$).

The observational model is a multinomial model with parameters the vector of probabilities $\bm{p}_{it}=(p_{it,1},\dots,p_{it,K})$, where $p_{it,k}$ is the probability of belonging to class $k$ per parcel $i$ and time $t$,
%
\begin{equation*}
y_{it} \sim \text{multinomial}(\bm{p}_{it}).
\end{equation*}

@@

In the case of the first-order components, the normalized lengthscale estimates $\left(\frac{2\cdot \hat{\ell}_{d,1}}{|x^d_{max}-x^d_{min}|}\right)$ are all bigger than the minimum lengthscale reported by Figure (6) % \ref{fig6_lscale_vs_J_vs_c} 
as a function of $m$, $c$. Which means that the used number of basis functions ($m=15$) and boundary factor ($c=2.5$) are suitable values for modeling accurately the input effects.

@@

The Stan model code for the approximate GP model of this case study can be found at \small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Land-use-classification}.

>>>>

\usepackage{natbib}
\bibliographystyle{plainnat}
\bibpunct{(}{)}{;}{a}{,}{,}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{float}
\usepackage{soul}
\usepackage{xcolor}
\graphicspath{{./images/}}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{url}
\usepackage{parskip}

\usepackage[titletoc]{appendix}
\usepackage{booktabs}
\usepackage{boldline}
\usepackage{colortbl}
\DeclareMathOperator{\LogNormal}{LogNormal}
\DeclareMathOperator{\GP}{\mathcal{GP}}
\DeclareMathOperator{\Normal}{Normal}

@@

% \bibpunct[, ]{[}{]}{,}{}{,}{,}
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}

\usepackage{listings}
\usepackage{slashbox}
\usepackage{caption}
\captionsetup{font=footnotesize}
\usepackage[colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black]{hyperref}

%\bibliographystyle{authordate1}
% TODO macro
\newcommand{\todo}[1]{\textcolor{blue}{\textbf{[#1]}}}
\newcommand{\resp}[1]{\textcolor{red}{\textbf{[#1]}}}

\title{Online supplemental material of the article:\\ \vspace{7mm} 'Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming'\\ \vspace{7mm}
\normalsize by}

\author{Gabriel Riutort-Mayol and Paul-Christian Bürkner and Michael R.\ Andersen and\\
  Arno Solin and Aki Vehtari}

@@

This document consists of four case studies of application of the article 'Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming'.

@@

\section{Same-sex marriage case study}\label{ch5_sec_studycaseII}
This data set relates the proportion of support for same-sex marriage to the age. The data consists of 74 observations of the amount of people $y_i$ supporting same-sex marriage from a population $n_i$ per age group $i$ ($i=1,\dots,74$). The observational model is a binomial model with parameters population $n_i$ and probability of supporting same-sex marriage $p_i$ per age group $i$,
%
\begin{equation*}
y_i \sim \mathrm{Binomial}(p_i, n_i).
\end{equation*}

\noindent The population per age group $n_i$ is a known quantity and the goal is to estimate the same-sex support probability $p_i$ or mean number of support people per age group. Probabilities $\bm{p}=(p_1,\dots,p_{74})$ are modeled by a Gaussian process (GP) function $f:{\rm I\!R} \to {\rm I\!R}$ with a squared exponential covariance function $k$, as a function of age input values $\bm{x}=(x_1,\dots,x_{74})$, and through the {\it logit} function as a link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \mathrm{logit}(f(x_i)) \nonumber \\
f(x) &\sim \mathcal{GP}(0, k(x,x', \theta)).
\end{align*}

\noindent Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,74$. The covariance function $k$ depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process.

In the HSGP model, the function $f(x)$ is approximated as in equation (7)%(\ref{eq_approxf})
, with the squared exponential spectral density as in equation (1)%(\ref{eq_specdens_inf})
, and eigenvalues $\lambda_j$  and eigenfunctions $\phi_j$ as in equations (5) %(\ref{eq_eigenvalue}) 
and (6)%(\ref{eq_eigenfunction})
.

In order to do model comparison, in addition to the regular GP model and HSGP model, an splines-based model is also fitted using the Thin Plate Regression Splines approach in \cite{wood2003thin} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this splines model using the R-package \textit{brms} \citep{burkner2017brms}.

@@

For the HSGP model, different models with different number of basis functions and boundary factor have been fitted. The root mean square errors (RMSE) for every one of these models have been computed against the regular GP model, and plotted as a function of the number of basis functions $m$ and boundary factor $c$ in Figure \ref{ch5_fig13_MSE_train_BF_gaydata}-left, for sample data, and in Figure \ref{ch5_fig13_MSE_train_BF_gaydata}-right, for test data. As the boundary factor increases, more basis functions are needed for a accurate approximation, as it was expected.

Figure \ref{ch5_fig14_MSE_train_gaydata} shows the RMSE of the regular GP, HSGP and splines models computed against the actual data, for training and test data, as a function of the number of basis functions $m$ and boundary factor $c$ for the HSGP model, and  knots for the splines model. We can see how the splines models do not extrapolate data properly (Figure \ref{ch5_fig14_MSE_train_gaydata}-right).

Figure \ref{ch5_fig14_1_time_gaydata} shows computational times, in seconds per iteration (iteration of the HMC sampling method), as a function of the number of basis functions $m$, for the HSGP model, and knots, for the splines model. The computational times is represented in the y-axis which is on a logarithmic scale. The HSGP model is on average roughly 15 times faster than the regular GP and 5 times faster than the spline model, for this particular case and univariate input space. Computation time increases relatively slight with the number of basis functions in a univariate input space, as can be seen in the figure.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Same-sex-marriage-data}}.

@@

This example consists of a simulated dataset with $n=120$ ($i=1,\dots,n$) single draws from a GP prior as function of input values $\bm{x}_i \in \{[-1,1],[-1,1]\} \subset {\rm I\!R}^2$ in a bivariate input space ($D=2$). A squared exponential covariance function, with hyperparameters marginal variance $\alpha=1$ and lengthscales $\ell_1=0.10$, for the first input dimension, and $\ell_2=0.35$, for the second input dimension, is used for the GP prior. Gaussian noise standard deviation $\sigma=0.2$ was added to the GP draws to form the final noisy set of observations $\bm{y} \in {\rm I\!R}^{120}$.

The regular GP model over the outcome variable $\bm{y}$ can be written as follows,
%
\begin{eqnarray*}\label{ch5_eq_latentgp_simudata2}
\begin{split}
\bm{y} &= \bm{f} + \bm{\epsilon} \\
\bm{\epsilon} &\sim \Normal(0, \sigma^2  I) \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x}, \bm{x}', \theta)),
\end{split}
\end{eqnarray*}

\noindent where $\bm{f}=\{f(\bm{x}_i)\}_{i=1}^{120}$ represents the underlying function at the input values $\bm{x_i}\in {\rm I\!R}^{2}$, $\bm{\epsilon}$ is the Gaussian noise term with variance $\sigma^2$, and $I$ represents the identity matrix. The function $f:{\rm I\!R}^2 \to {\rm I\!R}$ is a GP prior with a multivariate squared exponential covariance function $k(\bm{x}_i,\bm{x}_j,\theta)$, which depends on the inputs $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell_1,\ell_2\}$. The hyperparameters $\alpha$, $\ell_1$ and $\ell_2$ represent the marginal variance and lengthscales for first and second input dimensions, respectively, of the GP process. Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}$ is multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(\bm{x}_i,\bm{x}_j,\theta)$, with $i,j=1,\dots,120$.

The marginalized form, by integrating out the latent values $\bm{f}$, of the previous latent GP model results:
%
\begin{equation*}\label{ch5_eq_marginalizedgp_simudata2}
\bm{y} \sim \Normal(0, \bm{K} + \sigma^2 I ).
\end{equation*}

In the HSGP model with $2$ input dimensions, the latent function $f(\bm{x})$ is approximated as in equation (12)% (\ref{eq_approxf_multi})
, with the $2$-dimensional squared exponential spectral density $S$ as in equation (1)% (\ref{eq_specdens_inf})
, and the $2$-vector of eigenvalues $\bm{\lambda}_j$ and the multivariate eigenfunctions $\phi_j$ as in equations (9) %(\ref{eq_eigenfunction_multi})
 and (10)% (\ref{eq_eigenvalue_multi})
 , respectively.

In order to do model comparison, in addition to the regular GP and HSGP models, a two-dimensional splines-based model is also fitted using a cubic spline basis, penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized}, and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

Figure \ref{ch5_fig15_Posteriors_exII} shows the data-generating GP function, from where the dataset was drawn, and the mean posterior predictive functions of the three models, the regular GP, the HSGP, and the splines, fitted over the dataset. Sample observations are also plotted as circles. For the HSGP model, $m_1=40$ and $m_2=40$ basis functions for each dimension respectively, were used, which lead to a total of 1600 multivariate basis functions. A boundary factor for each dimension $c_1=1.5$ and $c_2=1.5$ were used. For the splines model, 40 knots in each dimension were used.

@@

In order to assess performance of the models as a function of the number of basis functions and knots, different models with different number of basis functions, for the HSGP model, and different number of knots, for the splines model, have been fitted. In all models, the same number of basis functions and knots per dimension were used. Figure \ref{ch5_fig17_RMSE_exII}-left shows the root mean squared error (RMSE), computed against the data-generating function, as a function of the boundary factor $c$, and the number of univariate basis functions $m$, for the HSGP model, and knots, for the splines model. From Figures \ref{ch5_fig16_errors_exII} and \ref{ch5_fig17_RMSE_exII}-left, it can be seen a close approximation of the HSGP model to the regular GP model. However, the performance of the splines model is significantly worse.

Figure \ref{ch5_fig17_RMSE_exII}-right shows the computational times of the different models as a function of the boundary factor, number of basis functions and knots. Figure \ref{ch5_fig17_RMSE_exII} reveals that choosing the optimal boundary factor allows for less number of basis functions and less computation time. Even though in a bivariate input space the computation demand increases significantly with the number of dimensions or knots, either the HSGP or spline models work significantly better than regular GP, even for highly wiggly functions that require a high number of basis functions or knots for a accurate approximation. However, serious difficulties with cumputation time were encountered in building the spline model with $50$ knots for this bivariate function.

The Stan model codes for the exact GP, the approximate GP and the splines models of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_2D-Simulated-data}}.

@@

\section{Diabetes case study}\label{ch5_sec_bf_caseV}
The next example presents an epidemiological study of diabetes disease. The study aims to relate the probability of suffering from diabetes to some risk factors. The data contains $n=392$ individuals ($i=1,\dots,n$) from which the binary variable of suffering ($y_i=1$) or not suffering ($y_i=0$) from diabetes have been observed. The input vector $\bm{x}_i=(x_{i1},x_{i2},x_{i3},x_{i4}) \in {\rm I\!R}^{4}$, in a 4D input space ($D=4$), contains the risk factors, \textit{Glucose} ($x_{i1}$), \textit{Pregnancy} ($x_{i2}$), \textit{Age} ($x_{i3}$) and \textit{BMI} ($x_{i4}$), per individual $i$. The observational model is a Bernoulli model with parameter the probability $p_i$ of suffering from diabetes per observation $i$,
%
\begin{equation*}
y_i \sim \mathrm{Bernoulli}(p_i).
\end{equation*}

\noindent The goal is to estimate the probability $p_i$ as a function of the risk factors, which function $f(\cdot):{\rm I\!R}^{4} \to {\rm I\!R}$ is modeled as a GP with a multivariate squared exponential covariance function $k$ depending on the risk factors $\bm{x}$ and hyperparameters $\theta=\{\alpha,\ell\}$, and related to the probabilities $p_i$ through the {\it logit} link function,
%
\begin{align*} \label{ch5_eq_gpprior_gay}
p_i &= \mathrm{logit}(f(\bm{x}_i)) \nonumber \\
f(\bm{x}) &\sim \mathcal{GP}(0, k(\bm{x},\bm{x}', \theta).
\end{align*}

\noindent Saying that the function $f(\cdot)$ follows a GP model is equivalent to say that $\bm{f}=\{f(\bm{x}_i)\}_{i=1}^n$ are multivariate Gaussian distributed with covariance matrix $\bm{K}$, where $K_{ij}=k(x_i,x_j,\theta)$, with $i,j=1,\dots,n$. The hyperparameters $\alpha$ and $\ell$ represent the marginal variance and lengthscale, respectively, of the GP process. A scalar lengthscale has been considered in the multivariate covariance function $k$.

@@

In order to do model comparison, in addition to the regular GP and HSGP models, a $D$-dimensional spline-based model is also fitted using a cubic spline basis penalized by the conventional integrated square second derivative cubic spline penalty \citep{wood2017generalized} and implemented in the R-package \textit{mgcv} \citep{wood2011mgcv}. A Bayesian approach is used to fit this spline model using the R-package \textit{brms} \citep{burkner2017brms}.

@@

Figure \ref{ch5_fig20_time2D_diabetes} shows the computational times of the different models, regular GP, HSGP and spline, fitted over the dataset with 2, 3 and 4 input dimensions, as a function of the boundary factor $c$ and number of univariate basis functions $m$ and knots. The time of omputation increases significantly with higher dimensions for the HSGP and spline models. In this sense, choosing optimal values for the number of basis functions and boundary factor, looking at the recommendations and diagnosis provided by Figure 6 % \ref{fig6_lscale_vs_J_vs_c}
in the main document, can be essential to avoid a excessive computation time, especially in multivariate input spaces. Serious difficulties with computation time were encountered in building spline models in a 4D input space or even in a 3D input space with more than 10 knots.

The Stan model codes for the exact GP, the approximate GP and the spline models of this case study can be found online at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Diabetes-data}}.

@@

The next example presents an spatio-temporal classification in land-use of plots between 2006 and 2015 in a part of the territory of Valencia in Spain dedicated to growing citrus fruits. A sampling set consists of $n=200$ plots with known class. The data is recorded in a time series of $T=5$ years (2006, 2008, 2010, 2012, 2015) within the period. The class of each parcel $i$ and time $t$ is stored by a categorical variable $y_{it}$ representing the $K=5$ different possible classes ($k=1,\dots,K$): $k=1$, adult independent citrus fruits; $k=2$, aligned citrus fruits; $k=3$, irregular citrus fruits; $k=4$, abandoned citrus fruits; $k=5$, young citrus fruits.

@@

The principal components (PCs) will be used jointly with the time variable as inputs in the classifying additive HSGP model. Let's denote the matrix $\bm{X}=[\bm{x}_{11}\, \cdots \, \bm{x}_{it} \, \cdots \, \bm{x}_{nT}]^\top \in {\rm I\!R}^{nT\times D}$, which contains the input vectors $\bm{x}_{it} \in {\rm I\!R}^D$, $D=53$ ($52 \text{ PCs plus time}$) for the spatio-temporal observations (plots $i=1,\dots,200$, and times $t=1,\dots,5$).

The observational model is a multinomial model with parameters the vector of probabilities $\bm{p}_{it}=(p_{it,1},p_{it,2},\dots,p_{it,K})$, where $p_{it,k}$ is the probability of belonging to class $k$ per parcel $i$ and time $t$,
%
\begin{equation*}
y_{it} \sim \mathrm{Multinomial}(\bm{p}_{it}).
\end{equation*}

@@

In the case of the first-order components, the normalized lengthscale estimates $\left(\frac{2\cdot \hat{\ell}_{d,1}}{|x^d_{max}-x^d_{min}|}\right)$ are all bigger than the minimum lengthscale reported by Figure (6) % \ref{fig6_lscale_vs_J_vs_c} 
as a function of $m$ and $c$. Which means that the used number of basis functions ($m=15$) and boundary factor ($c=2.5$) are suitable values for modeling accurately the input effects.

@@

The Stan model code for the approximate GP model of this case study can be found at {\small \url{https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper/Case-study_Land-use-classification}}.
